Functional Pearl --- Programming with \\ Negative, Fractional, Square Root, and Imaginary Types

### 2012-02-24 dual.tex
added some long term thoughts

>>>>

The conventional idea is to divide the world into a ``real'' one and a
``virtual'' one. In the ``real'' world, we can define datatypes like
\verb|t=t^2+1| but we don't have additive inverses so it makes no sense to
talk of negative types and we can't rearrange the terms in the datatype
definition. However the observation is that we can map these datatypes to a
virtual world that has more structure (a ring that provides additive inverses
or a field that also provides multiplicative inverses) and then perform
computations in the ring/field. If we perform computations in the ring, then
some of these will use additive inverses in ways that cannot be mapped back
to the ``real'' world. Much of current research attempts to characterize
which computations done in the ring are valid isomorphisms between datatypes
in the ``real'' world. This is nice but is not what I am after. In fact I am
not interested in the ring or the semiring at all. I am interested in the
field and I want this field to be \textbf{the real world.} This is partly
motivated by the fact that Quantum Mechanics seems to demand an underlying
field and more generally that the field provides the maximum generality in
slicing and dicing computations. So assuming I live in a field and that the
negative, fractional, square root, and imaginary types are all ``real,'' how
do I compute in this field? Clearly there will be constraints on
``measurement'' in the sense that a full program cannot produce any of the
crazy types but that's done outside the formalism in some sense just as in
Quantum Computing. The main question I am after is how to compute in this
field with first-class negative, fractional, etc. types. As I mentioned in my
previous email, we can produce programs that have types \verb|t^3 <-> -1| and
they ``run'' (but only to give infinite loops).

So when a programmer writes the datatype declaration \verb|t = t^2+1|, if we
allow negative etc. then this is effectively writing 
\verb|t = cubicroot{-1}|. If we are in the field then computations that 
manipulate these trees can be sliced and diced even at interfaces that
expose the cubic root and the imaginary types.

Future work: develop a type system for a ``normal language'' that has
negative, fractional, etc. types as first-class types. More long term,
instead of adding one polynomial at a time, we can go to an algebraically
closed field. The complex numbers is an obvious choice but I would rather go
to something computable like the field of algebraic numbers. Is the adele
ring or the p-adics relevant here?
